---
title: "Final Project"
author: "jvogel, smillard, tsmith"
date: "2022-11-05"
output:
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(ggplot2)
library(dplyr)
library(outliers)
options(dplyr.summarise.inform = FALSE)
```

Getting dataset from R:

https://stat.ethz.ch/R-manual/R-devel/library/datasets/html/esoph.html

Setting up our data

```{r}
dat <- esoph

#adding column to data of cancer percentages

esoph$percentage <- esoph$ncases / (esoph$ncontrols+esoph$ncases)

```

Prove Data is "clean", show no outliers, using cooks distance test.

```{r}

#Here, we want to ensure that all of the data is clean. The techniques can help spot irregularities in the data.

#summary(dat)
boxplot(dat$agegp,dat$alcgp, dat$tobgp)

model <- lm(percentage ~ agegp+ alcgp+tobgp, data = dat)
summary(model)


#The data is already pretty clean and ready to use.
```



```{r}
#Here, we try the Shapiro-Wilk test with probability.

model2 <- aov(percentage ~ agegp + tobgp + alcgp, data = esoph) #A Linear model is created in order to apply ANOVA test.
plot(model2$residuals)
hist((model2$residuals))
#This is the formal test of normality.
shapiro.test(model2$residuals)
```

**We can conclude that by the p-value being > .05,  there is is not enough statistical evidence to reject the null hypotheses that the data is normally distributed. The histogram here reflects a normal distribution.**

-----------------

**In this next section, we perform a  three-way ANOVA test which we used to determine whether or not there is a statistically significant difference between the means of our independent groups. Our independent groups are age, alcohol consumption, and tobacco consumption. Each of our independent groups have several levels: age group and consumption amounts.**



----------------



```{r}
#The three-way anova has three independent variables and one dependent variable.
esoph.aov3way <- aov(percentage ~ alcgp + tobgp + agegp, data = esoph)
#Here we use the ANOVA
anova(esoph.aov3way)
#summary(esoph.aov3way)
```

**We see from the p-values that there is enough statistical evidence to show that we reject the null hypothesis that there is no difference in means for alcohol consumption and age group. But there is not enough evidence to reject the null hypothesis that there is no difference in means for tabacoo consumption**

**To follow up with our ANOVA test, we performed a Tukey's test on our significant factors Alcohol consumption and age group to find which pairs of levels are significant.**

**Here, the Tukey's test is being performed.**

```{r}
#Here, we perform a Tukey's test, on the alcohol
tukey.alcohol <- TukeyHSD(esoph.aov3way, which = "alcgp" )
tukey.alcohol


#Here, we plot the CI for the Tukey's test on the "age" variable.
plot(tukey.age)
```
**Using both CI and p-values, the Tukey's test results for the "alcohol consumption" variable reflects that the pairs that are significant are (p-value <. 05): (80-119) & (0-39) g/day; (120+) & (0-39) g/day; (120+) & (40-79) g/day; (120+) & (80-119) g/day** 

```{r}
#Here, we display the results of Tukey's tests.

tukey.age <- TukeyHSD(esoph.aov3way, which = "agegp" )
tukey.age

#Here, we reflect the plot for the Tukey's test on the "alcohol consumption" variable.
plot(tukey.alcohol)

```

**Using both CI and p-values, the Tukey's test results for the "age" variable reflects that the pairs that are significant are (p-value < .05): (45-54) & (25-34); (55-64) & (25-34); (65-74) & (25-34); (75+) & (25-34); (45-54) & (35-44); (55-64) & (35-44); (65-74) & (35-44); (75+) & (35-44)**
