---
title: "Final Project"
author: "jvogel, smillard, tsmith"
date: "2022-11-05"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(ggplot2)
library(dplyr)
library(outliers)
options(dplyr.summarise.inform = FALSE)
```

Getting dataset from R:

https://stat.ethz.ch/R-manual/R-devel/library/datasets/html/esoph.html

```{r}
dat <- esoph
```

Prove Data is "clean", show no outliers, cooks distance test? 

```{r}
#Here, we want to ensure that all of the data is clean. The techniques can help spot irregularities in the data.
#head(dat)
#summary(dat)
#boxplot(dat)

chisq.out.test(dat)

#model <- lm(agegp ~ alcgp, data = outliers)

#cooksD <- cooks.distance(model)

#The data is already pretty clean and ready to use.
```

**First, we model our data in order to check if the residuals are normally distributed in order to conduct the ANOVA.**

```{r}
model <- aov(dat$ncases ~ dat$agegp+dat$alcgp+dat$tobgp, data = dat)

#Summarizing the model
summary(model)

#Plotting the model
plot(model)

#Looking at histogram of residuals
hist(model$residuals, main = "Residuals Histogram")
```
  


```{r}
shapiro.test(model$residuals)
```

**From this Shapiro-Wilk normality test, we can tell that the data is not normally distributed.**


```{r}
#Here, we try the Shapiro-Wilk test with probability.
esoph$percentage <- esoph$ncases / (esoph$ncontrols+esoph$ncases)

model2 <- aov(percentage ~ agegp + tobgp + alcgp, data = esoph) #A Linear model is created in order to apply ANOVA test.

plot(model2$residuals)
hist((model2$residuals))


#This is the formal test of normality.
shapiro.test(model2$residuals)
```

**We can conclude that by the p-value being > .05, we will not reject the null as there is strong evidence against rejecting the null. The histogram here reflects a normal distribution.**

-----------------

**In this next section, we perform a two-way and three-way ANOVA test which we used to determine whether or not there is a statistically significant difference between the means of our independent groups. Our independent groups are age, alcohol consumption, and tobacco consumption. Each of our independent groups have several levels: age group and consumption amounts.**

**To follow up with our ANOVA test, we performed a Tukey's test on Alcohol consumption and age group to find which pairs are significant.**

----------------

**First lets try the two-way ANOVA:**

```{r} 
#We need our dependent variable (percent of cancer cases) and two independent variables (tobacco and alcohol use) in their own data set.
esoph.twoway <- esoph[ , c("alcgp", "tobgp", "percentage")]
```


```{r}
#Next, a frequency table is generated.
table(esoph.twoway$alcgp, esoph.twoway$tobgp)

#Here, the table shows that we have a relatively balanced design.
```


```{r}
#We create the anova test
esoph.aov2way <- aov(percentage ~ alcgp + tobgp, data = esoph.twoway)

#Then use summary to summarize the test
summary(esoph.aov2way)

#The model above is called an additive model hence the + sign, but if we think the variables interact we can use an * instead of the +
```


```{r}
#The three-way anova has two independent variables and one dependent variable.
esoph.aov3way <- aov(percentage ~ alcgp + tobgp + agegp, data = esoph)

#Here we use summary to summarize the test
summary(esoph.aov3way)
```


**Here, the Tukey's test is being performed.**

```{r}
#Here, we perform a Tukey's test, and assign them to variables for plots later.
tukey.alcohol <- TukeyHSD(esoph.aov3way, which = "alcgp" )
tukey.age <- TukeyHSD(esoph.aov3way, which = "agegp" )
```



```{r}
#Here, we display the results of Tukey's tests.
tukey.age
tukey.alcohol
```
**Using both CI and p-values, the Tukey's test results for the "age" variable reflects that the pairs that are significant are (p-value < .05): (45-54) & (25-34); (55-64) & (25-34); (65-74) & (25-34); (75+) & (25-34); (45-54) & (35-44); (55-64) & (35-44); (65-74) & (35-44); (75+) & (35-44)**

**Using both CI and p-values, the Tukey's test results for the "alcohol consumption" variable reflects that the pairs that are significant are (p-value <. 05): (80-119) & (0-39) g/day; (120+) & (0-39) g/day; (120+) & (40-79) g/day; (120+) & (80-119) g/day** 


```{r}
#Here, we reflect the plot for the Tukey's test on the "age" variable.
plot(tukey.age)
```


```{r}
#Here, we reflect the plot for the Tukey's test on the "alcohol consumption" variable.
plot(tukey.alcohol)
```